Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-19 latex/main.tex
Add counts for parsimony
QC section pass

Given a newly attached node, mutation collapsing inspects its siblings from
previous sample days to check if any of them share (a subset of) the mutations
that
it carries. If so, we increase the overall parsimony of the inference
by creating a new node representing the ancestor that carried
those shared mutations and make that new node the parent of the
siblings carrying those shared mutations.
The patterns of
shared mutations between siblings can be complex, and the current
implementation uses a simple greedy strategy for choosing
the particular mutations to collapse.

The reversion push operation inspects a newly added node to see
if any of its mutations are ``immediate reversions''; that is,
are reversions of a mutation that occurred on the new node's
immediate parent. We increase the overall parsimony of the
inference by ``pushing in'' a new node which descends from the
original parent, and carries all its mutations except those
causing the reversions on the newly added node.

@@

\subsubsection*{ARG postprocessing}
The first step in ARG postprocessing is to examine putative recombinants to
determine whether a more parsimonious explanation is possible. This is done 
by rematching the inferred recombinant sequence against the ARG pruned back
to include only samples before the date of the recombinant. Additionally,
 each recombination node is examined to determine if any mutations 
are shared [TODO Yan write this up].
We refer to this process as ``long branch splitting''. 
% Data from
% https://github.com/jeromekelleher/sc2ts-paper/blob/main/arg_postprocessing/logs/sc2ts_v1_2023-02-21_rewire-recombinants-report.ipynb
From this, 58 recombinants were identified as being more parsimoniously 
explained by matching to existing nodes in the ARG (inserted by 
subsequent parsimony improving heuristics) and 16 by long branch 
splitting. Of these 16 recombinants, two are noteworthy as 
being the origins of the Delta and BA.2 lineages in the ARG.
Both result in substantially fewer mutations (15 fewer for 
Delta, 6 for BA.2) than the previous recombinant HMM solution.
These major saltational lineages highlight the challenges 
of using the LS HMM with highly diverged samples.
We then updated the ARG to remove these 74 artefactual 
recombinants by ``rewiring'' the recombination node to the 
more parsimonious non-recombinant solution. This 
resulted in 16 additional nodes, 182 fewer mutations and reduced
the number of trees along the genome from 348 to 317.

@@

\subsubsection*{Quality control of recombinants}

@@

Before characterizing recombination events in detail,
we sought to remove probable artifactual recombinants
due to correlated bases, which can be caused by
bioinformatics errors in genome assembly \cite{Turakhia2020},
multi-nucleotide substitutions (MNS) \cite{DeMaio2024}, and
alignment errors \cite{Hunt2024}).

We characterized artifactual recombination events
as having fewer than four such loci supporting the left or right parent
on either side of a suggested breakpoint.
We treated adjacent and near-adjacent sites
(separated by at most three bases) as single loci.
Also, when counting loci supporting a focal parent,
we penalized loci where the character(s)
in the recombinant matches that in the non-focal parent
but not in the focal parent.

Of the 929 recombination events in the ARG,
this measure identifies 543 as artifactual,
leaving 354 QC-passing events (Figure~\ref{fig:recombinant_qc}).

Furthermore,
the example copying patterns in Figure~\ref{fig:copying_pattern_examples}
illustrate different types of artefacts removed using this filter.
Q1 shows an example of a recombinant caused by an MNS (A507T, T508C, G509A),
which is a documented sequencing error due to incomplete read trimming \cite{DeMaio2024}.
Q4 shows an example of a recombinant caused by a Delta lineage-defining deletion.
The vast majority (XX, or XX\%) of the QC-failing recombinants has this deletion, and
many cases are affected by a known amplicon dropout caused by this particular deletion.
Q3 shows an example artefact caused by a combination of an MNS and a deletion.

Evidence that this filter effectively removed artifacts can be seen
in the clear association between the AmpliSeq V1 kit and
recombination events that we identified as artifactual.

>>>>

Given a newly attached node, mutation collapsing inspects its siblings from
previous sample days to check if any of them share (a subset of) the mutations
that
it carries. If so, we increase the overall parsimony of the inference
by creating a new node representing the ancestor that carried
those shared mutations and make that new node the parent of the
siblings carrying those shared mutations.
The patterns of
shared mutations between siblings can be complex, and the current
implementation uses a simple greedy strategy for choosing
the particular mutations to collapse. 
119,844 nodes were 
added to the ARG during primary inference by mutation collapsing.

The reversion push operation inspects a newly added node to see
if any of its mutations are ``immediate reversions''; that is,
are reversions of a mutation that occurred on the new node's
immediate parent. We increase the overall parsimony of the
inference by ``pushing in'' a new node which descends from the
original parent, and carries all its mutations except those
causing the reversions on the newly added node.
52,336 nodes were added to the ARG during primary inference by
reversion pushing.

@@

\subsubsection*{ARG postprocessing}
The first step in ARG postprocessing is to examine putative recombinants to
determine whether a more parsimonious explanation is possible. This is done 
by rematching the inferred recombinant sequence against the ARG pruned back
to include only samples before the date of the recombinant.

Where the recent ancestry of a recombinant includes a branch containing
large numbers of mutations, it is possible
that a non-recombinant solution could be found by postulating an intermediate
node along that branch, that contains only a subset of those mutations.
In particular, for a given mutation-rich branch, artefactual recombinants may
be found in which one recombinant region matches to
a node like the parent whereas another region matches to a node like the child
of that branch. Such recombinants can be identified by the
presence of recurrent or reversion mutations in the left or right parent branches.
To help remove these recombination events from the ARG, we note the genomic positions
of mutations associated with the recombinant, and identify branches
in their ancestry which have mutations at the same positions. Of these branches,
we chose the one that contains the largest total number of mutations as
a candidate for intermediate node creation. This is done by sorting the
mutations on the branch such that the ones that match the allelic state in the
recombinant are placed above the intermediate node, and the others are left
below the node. We refer to this process as ``long branch splitting'', and it
creates additional nodes which are available for recombinant rematching.

% Data from
% https://github.com/jeromekelleher/sc2ts-paper/blob/main/arg_postprocessing/logs/sc2ts_v1_2023-02-21_rewire-recombinants-report.ipynb
Using recombinant-rematching, 58 recombinants were identified as being more parsimoniously 
explained by matching to existing nodes in the ARG (inserted by 
subsequent parsimony improving heuristics) and 16 by long branch 
splitting. Of these 16 recombinants, two are noteworthy as 
being the origins of the Delta and BA.2 lineages in the ARG.
Both result in substantially fewer mutations (15 fewer for 
Delta, 6 for BA.2) than the previous recombinant HMM solution.
These major saltational lineages highlight the challenges 
of using the LS HMM with highly diverged samples.
We then updated the ARG to remove these 74 artefactual 
recombinants by ``rewiring'' the recombination node to the 
more parsimonious non-recombinant solution. This 
resulted in 16 additional nodes, 182 fewer mutations and reduced
the number of trees along the genome from 348 to 317. The long branches
involved in the origins of Delta and BA.2 originally joined
B.1 to B.1.167 and B.1.1 to BA.1 respectively; they can be seen
in Figure \ref{fig:subgraph_delta} and \ref{fig:subgraph_ba1_ba2}
with inserted intermediate nodes leading to the Delta and BA.2
origination nodes.

@@

\subsubsection*{Quality control of recombination events}

@@

Sequencing errors in genome assemblies \cite{Turakhia2020},
read misalignments \cite{Hunt2024}, and
multi-nucleotide substitutions (MNS \cite{DeMaio2020issues})
can cause short regions of a sample genome to match more closely
to an alternative genome rather than the true parent.
Such genomic regions can be mistaken for evidence of recombination,
because the LS HMM counts each base separately.
Fortunately, artefactual recombinants caused by this effect can be identified
because they affect multiple adjacent or near-adjacent bases simultaneously.

We devised a quality control (QC) measure based on this observation, and
applied it to each putative recombination event in the ARG.
We listed the genomic positions at which the suggested parents differed in allelic state.
Where such sites clustered closely (3 or fewer bases apart),
we treated the entire cluster of sites as a single \emph{locus}.
For each genomic region identified by the HMM as having a single parent,
we calculated the "net number of supporting loci" by counting the number of loci in the region
that supported the suggested parent minus the number that did not support it.
Finally, with all the recombination events in the ARG having a single breakpoint,
we calculated the net number of supporting loci on the left and right sides of each breakpoint, and
used this to plot the 855 recombination events in Figure~\ref{fig:recombinant_qc}.

This simple QC measure effectively differentiates
artefactual recombination events from those that are well supported.
All the sc2ts-detected recombination events associated with the Pango X lineages (section XX)
occur in the upper right of Figure~\ref{fig:recombinant_qc},
with high numbers of net supporting loci on both the right and left parents.
In contrast, the other regions of the plot, where one or both parents are less well supported,
are dominated by recombination events associated with samples sequenced using a specific protocol (AmpliSeq).
Coupled with closer inspection of copying patterns (below),
we chose a QC cutoff of four or more net supporting loci on both the left and right side of the breakpoint.
This leads to 354 QC-passing recombination events (Figure~\ref{fig:recombinant_qc}, upper right quadrant),
compared to 501 recombination events for which there is less support and which may be artefactual (shaded quadrants).

Inspection of the copying patterns for all the recombination events
clearly identified associations between recombination events
that failed this QC check and known problematic loci in the SARS-CoV-2 genome.
Figure~\ref{fig:copying_pattern_examples} illustrates examples of these copying patterns (Q1 to Q4),
in which adjacent sites are underlined in red, and near-adjacent sites (within 3 bases) in orange.
A continuous red+orange underline thus indicates a region treated as a single locus.
Q1 shows a well supported recombination event that passed QC (albeit marginally),
having a net number of supporting loci of 4 on both sides of the breakpoint.
Q2 shows an artefactual recombination event caused by a 9-base Delta lineage-defining deletion.
There are XX (XX\%) of the recombination events involving this deletion that failed QC, and
many cases are affected by a known amplicon dropout caused by this particular deletion.
Q3 shows an artefactual recombination event caused by a combination of
this Delta lineage-defining deletion and a known MNS (see below).
Finally, Q4 shows an artifactual recombination event associated with the A507T-T508C-G509A MNS,
which has been identified as a sequencing error due to incomplete read trimming \cite{DeMaio2024}.
There are XX recombination events involving this particular MNS, of which XX\% failed QC.
